Computational Prediction of Boundaries of 3D Genomic Domains in Class Imbalance

### 2019-04-07 Manuscript/manuscript.Rmd
manuscript edits

Multiple random forest classification algorithms were built on domain data extracted from Hi-C data for 2 cell lines $C = \{GM12878, K562\}$, at four different resolutions $R = \{\text{10 kb}, \text{25 kb}, \text{50 kb}, \text{100 kb\}}$. For each classifier, we compared three predictor types $P = \{\text{OC}, \text{OP}, \text{Distance}\}$ and 4 class balancing techniques $B = \{None, RUS, ROS, SMOTE\}$. A system produced from the ensemble framework was defined as: $S = \{c, r, p, b\}$, where $c \in C$, $r \in R$, $p \in P$, and $b \in B$. Letting $|X|$ denote the cardinality of each set, we evaluated $|C| \times |R| \times |P| \times |B|$ different ensemble systems, totaling $96 (= 2 \times 4 \times 3 \times 4)$ separate models.

@@

We examined the class imbalance of the different data resolutions for each cell line. As expected, the total number of genomic bins decreased as resolution decreased (Figure 4A). That is, at lower resolutions, the linear genome was binned into larger units, thereby creating fewer bins. The imbalance ratios, however, remained relatively stable across resolutions (Figure 4B). We found that there was an average IR of 0.154 (SD=0.022) and 0.137 (SD=0.0181) across resolutions for the GM12878 and K562 cell lines respectively. Ideally, a balanced dataset will have an IR close to 0.50. Here, the low imbalance ratios indicated a wide disparity between majority and

@@

We first evaluated performances for models implemented using overlap counts predictor types, without re-sampling (on the GM12878 cell line), as these characteristics represented models with the least granularity and could act as a baseline. In evaluating such composite metrics like F1-Score and MCC, we found that these models exhibited poor performances across resolutions (Figure 5A). Furthermore, at lower resolutions (50 kb and 100 kb specifically), the F1-Score and MCC could not be determined given by the model's complete inability to correctly classify any genomic bins that overlapped with flanked TAD boundaries in the test data, thereby resulting in 0 true positives, as can be seen by the missing red points. When including overlap percents perdictor types in the models, coupled with no resampling, we found that performances were worse than when using OC predictor types (Figure 5B). As with OC predictor types, these models were also not able to yield F1-Scores and MCCs at the 50 kb and 100 kb resolution. Models built on distance-type predictors, however, exhibited slightly better performances than the other two types, yet still yielded missing or low F1-Scores and MCC at lower resolutions (Figure 5C). This indicated that the key driver leading to poor performances was not feature engineering, but instead the severity in the class imbalance.

@@

Comparisons of the effects of the various resampling techniques were then made. We found that there was a substantial increase in F1-Score and MCC for models built on resampled data using any of the 3 techniques (Figure 5). After resampling, we were able to report performances at the lower resolutions which was not feasible for models built on fully imbalanced data. For all predictor types, implementation of either RUS or SMOTE resampling was found to yield consistently better results than no resampling and ROS.

@@

We then assessed the performances of models implemented on distance type predictors alone. Results showed that models with distance-type predictors outperformed models with other predictor types (Figure 5C). This result was consistent across both resampling technique and resolution. Directly comparing model performances using distance type predictors, we found that SMOTE resampled datasets yielded the highest model performances across each of the composite metrics. These results were consistent across resolution and cell line. Although RUS performed only marginally worse than SMOTE, there was a steap drop off in performances for ROS and no re-sampling, across both resolution and cell line. These results imply that a best performing model, one that is able to best discriminate between genomic bins that are near versus far away from TAD boundaries, is one that incorporates both random under- and over-sampling, as SMOTE is known to do.

>>>>

Multiple random forest classification algorithms were built on domain data extracted from Hi-C data for 2 cell lines $C = \{GM12878, K562\}$, at four different resolutions $R = \{\text{10 kb}, \text{25 kb}, \text{50 kb}, \text{100 kb\}}$. For each classifier, we compared model performances using 3 separate predictor types $P = \{\text{OC}, \text{OP}, \text{Distance}\}$ and 4 class balancing techniques $B = \{None, RUS, ROS, SMOTE\}$. A system produced from the ensemble framework was defined as: $S = \{c, r, p, b\}$, where $c \in C$, $r \in R$, $p \in P$, and $b \in B$. Letting $|X|$ denote the cardinality of each set, we evaluated $|C| \times |R| \times |P| \times |B|$ different ensemble systems, totaling $96 (= 2 \times 4 \times 3 \times 4)$ separate models.

@@

We examined the class imbalance of the different data resolutions for each cell line. As expected, the total number of genomic bins decreased as resolution decreased (Figure 4A). Specifically, at lower resolutions, the linear genome was binned into larger units, thereby creating fewer bins. The imbalance ratios, however, remained relatively stable and consistent across resolutions and cell line (Figure 4B). We found that there was an average IR of 0.154 (SD=0.022) and 0.137 (SD=0.0181) across resolutions for the GM12878 and K562 cell lines respectively. Here, the low imbalance ratios indicated a wide disparity between the number of genomic bins that overlap with a flanked TAD boundary and the much larger number that did not.

@@

We first evaluated performances for models implemented using OC predictor types, without re-sampling (on the GM12878 cell line), as these characteristics represented models with the least granularity and could act as a baseline. In evaluating such composite metrics like F1-Score and MCC, we found that these models exhibited poor performances across resolutions (Figure 5A). Furthermore, at lower resolutions (50 kb and 100 kb specifically), the F1-Score and MCC could not be determined given by the model's complete inability to correctly classify any genomic bins that overlapped with flanked TAD boundaries in the test data, thereby resulting in 0 true positives <!--, as can be seen by the missing red points-->. When including OP perdictor types in the models, coupled with no resampling, we found that performances were worse than when using OC predictor types (Figure 5B). As with OC predictor types, these models were also not able to yield F1-Scores and MCCs at the 50 kb and 100 kb resolution. Models built on distance-type predictors, however, exhibited slightly better performances than the other two types, yet still yielded missing or low F1-Scores and MCC at lower resolutions (Figure 5C). Similar results were obtained on the K562 cell line, in which F1-Scores and MCCs were either low or missing across resolutions, with marginal improvement when considering distance types (Supplementary Figure 10 A-C). This indicated that the key driver leading to poor performances was not due to feature engineering, but instead the severity in the class imbalance.

@@

Comparisons of the effects of the various resampling techniques were then made. On the GM12878 cell line, we found that there was a substantial increase in F1-Score and MCC for models built on resampled data using any of the 3 techniques (Figure 5). After resampling, we were able to report performances at the lower resolutions which was not feasible for models built on fully imbalanced data. For all predictor types, implementation of either RUS or SMOTE resampling was found to yield consistently better results than no resampling and ROS. These results were confirmed when evaluating the K562 cell line (Supplementary Figure 10 A-C).

@@

On the GM12878 cell line, results showed that models with distance-type predictors outperformed models with other predictor types (Figure 5C). This result was consistent across both resampling technique and resolution. Directly comparing model performances using distance type predictors, we found that SMOTE resampled datasets yielded the highest model performances across each of the composite metrics. Although RUS performed only marginally worse than SMOTE, there was a steap drop off in performances for ROS and no re-sampling, across resolution. The same combination of resampling technique and predictor type was also found to result in the best performing model for the K562 cell line (Supplementary Figure 10C). These results imply that a best performing model, one that is able to best discriminate between genomic bins that are near versus far away from TAD boundaries, is one that incorporates both random under- and over-sampling, as SMOTE is known to do.
